Modeling the Java Card API in B

### 2012-04-02 trunk/papers/B2012-work/javacardAPIpaper.tex
mais texto sobre a modelagem e ajustes nas referencias.

\begin{document}
\begin{frontmatter}
%  \title{Modeling the Java Card API (or Interface?) in B} \author{A. Martins Moreira\thanksref{ALL}\thanksref{anamaria}}
%  \address{DIMAp-UFRN, Natal, RN, Brazil} \author{S. de Oliveira Santos \thanksref{simone}}
%  \address{DIMAp-UFRN, Natal, RN, Brazil} \author{B. Emerson Gurgel Gomes \thanksref{bruno}}
%    \address{IFRN, Currais Novos, RN, Brazil} \author{D. Boris Paul D\'{e}harbe \thanksref{david}}
%      \address{DIMAp-UFRN, Natal, RN, Brazil} \thanks[ALL]{INES-CAPES-projeto GD?... o que mais?} \thanks[anamaria]{Email:
%    \href{mailto:anamaria@dimap.ufrn.br} {\texttt{\normalshape
%        anamaria@dimap.ufrn.br}}} 
%      \thanks[simone]{Email:
%    \href{mailto:simone82@ppgsc.ufrn.br} {\texttt{\normalshape
%        simone82@ppgsc.ufrn.br}}}
%      \thanks[bruno]{Email:
%    \href{mailto:bruno.gurgel@ifrn.edu.br} {\texttt{\normalshape
%        bruno.gurgel@ifrn.edu.br}}}
%       \thanks[david]{Email:
%    \href{mailto:david@dimap.ufrn.br} {\texttt{\normalshape
%        david@dimap.ufrn.br}}}

  \title{Modeling the Java Card API (or Interface?) in B}
  \author[ufrn]{A. Martins Moreira\thanksref{ALL}\thanksref{anamaria}}
  \author[ufrn]{S. de Oliveira Santos \thanksref{simone}}
  \author[ifrn]{B. Gurgel Gomes \thanksref{bruno}} \author[ufrn]{D.
    D\'{e}harbe \thanksref{david}} \address[ufrn]{DIMAp-UFRN, Natal,
    RN, Brazil} \thanks[ALL]{This work is partly supported by CAPES
    and CNPq grants 306033/2009-7, 560014/2010-4 and 573964/2008-4
    (National Institute of Science and Technology for Software
    Engineering---INES, \texttt{www.ines.org.br}).}
\address[ifrn]{IFRN, Currais Novos, RN, Brazil}
\thanks[anamaria]{Email: \href{mailto:anamaria@dimap.ufrn.br}
  {\texttt{\normalshape anamaria@dimap.ufrn.br}}}
\thanks[simone]{Email: \href{mailto:simone82@ppgsc.ufrn.br}
  {\texttt{\normalshape simone82@ppgsc.ufrn.br}}}
\thanks[bruno]{Email: \href{mailto:bruno.gurgel@ifrn.edu.br}
  {\texttt{\normalshape bruno.gurgel@ifrn.edu.br}}}
\thanks[david]{Email: \href{mailto:david@dimap.ufrn.br}
  {\texttt{\normalshape david@dimap.ufrn.br}}}

@@

Among the different components of the Java Card platform, the Java
Card API is the one without which no refinement into Java Card code
could be done.  Suppose, for instance, that we had the B abstract
machine of Fig.~\ref{lis:buscard}, which models a bus card with a
variable that keeps track of the current credit the user has
(\texttt{balance}) and an operation to add credits to this balance
(\texttt{add}). This simple example provides enough evidence of the
importance of having a B model of the Java Card API.  As all data in
it is implementable in C, for instance, and as it is a deterministic
specification, B tools can directly generate code for it in C, and the
resulting code would be very similar to the original B machine.

\lstset{language=B, caption={B machine Buscard},label=lis:buscard}
\begin{lstlisting}
MACHINE
	Buscard
CONSTANTS
	maximum
PROPERTIES
	maximum : INTEGER & MAX_BALANCE  = 100
VARIABLES
	balance
INVARIANT
	balance : INTEGER & balance <= MAX_BALANCE 
INITIALISATION
	balance := 0
OPERATIONS
	add(credit) =
	PRE
		credit : INTEGER & credit > 0  & 
                balance + credit <= MAX_BALANCE 
	THEN
		balance := balance + credit
	END
END
\end{lstlisting}

@@

\lstset{language=Java, caption={Method add in applet Java Card},label=lis:addjavacard}
\begin{lstlisting}
public void add (APDU apdu) { 
	byte[] bf;  short cr; byte apdu_state; short res;

@@

		if (!((short) (balance + cr) <= MAX_BALANCE)) {
			ISOException.throwIt(EX_BALANCE_EXCEEDED);
		} else if (!((short) (balance + cr) >= 0)) {
			ISOException.throwIt(EX_NEGATIVE_BALANCE);
		} else if (! (cr > 0)) {
			ISOException.throwIt(EX_NEGATIVE_CREDIT);
		} else {
			balance = (short) (balance + cr);
		}
	} 
}
\end{lstlisting}

@@

		IF not (sum_short (balance, cr) <= MAX_BALANCE)
		THEN 
			ISOException.throwIt(EX_BALANCE_EXCEEDED)
		ELSIF not (sum_short (balance, cr) >= 0)
		THEN ISOException.throwIt(EX_NEGATIVE_BALANCE)
		ELSIF not (cr > 0)
		THEN ISOException.throwIt(EX_NEGATIVE_CREDIT) 
		ELSE  
			balance := sum_short(balance, cr)
		END
	END
END
\end{lstlisting}

@@

The change of interface is also taken care of by the BSmart method, so
that the development can be done without need to extend the B
method, as detailed in~\cite{Gomes10}.

But all of these transformations rely on the need to reference Java
Card API types, classes and methods on the B level.  This is then the
reason why the Java Card API needed to be the first Java Card
component to be modeled in B.

[[[(PARA SIMONE e BRUNO: maybe also cite related work ? Which are the formalization works that did not focus on the API?)]]]

[[[ PARA SIMONE -> tambem precisamos com urgencia daquela analise sobre o que muda da 2.2.2 para a 3 ]]]

@@

@@

\item The hierarchy of the Java Card API objects is partially modeled
  by abstract set inclusion. For instance, the class Throwable extends
  the class Object, in Java Card. In our model, we also have that the
  machine Throwable EXTENDS the machine Object, but the meaning of a
  machine extension in B is different from the meaning of a class
  extension in Java Card. To show that all throwables are indeed
  objects, a property stating that THROWABLE is a subset of OBJECT is
  included in the Throwable machine.

@@

As previously stated, we chose to model exception raising conditions
as operation pre-conditions in the B models. For instance, if the Java
Card specification states that a method should throw an exception in
case a parameter array is empty, the pre-condition of the
corresponding B operation will require that this parameter is not
empty. The B method then requires that every call to that operation
satisfies the operation's pre-condition. As a consequence, code that
is formally generated using the B method and the Java Card API
specification should never raise this specific exception, as it is
statically proved that all calls to this operation are made with
non-empty parameter arrays. Not all situations leading to exceptional
conditions are modeled, however, as some of them depend on the Java
Card Runtime Environment (JCRE) and its state, which has not yet been
modeled, and some require a more detailed specification of the
contents of some data and other internal behavior. In these cases, the
exception raising conditions are only left as comments to the B
specification, and can be used as additional documentation or to be
considered in future work.

One delicate issue to be dealt with both by the user of the API
specification and the tools designed to work with it, is the use of
overloaded methods in an application development. When overloaded
operations are used in an application development, the user must
explicitly choose which version of the overloaded operation is to be
used. Then, verifications will be carried out considering this
specific version of the operation. When code is generated, either the
code generator will remove the name extension automatically, if
designed for it, or the user will have to edit the generated code
manually. In any case, it is important that the actual method called
in the Java Card code is the one with which verifications have been
carried out.  In case of overriding, in particular, it will be either
the user's responsibility to choose the B one which is in scope at
this specific point in Java, or an extra verification step will be
included in the code generator to check that the operation used in
B is the one in scope in Java and registers an error in case it is
not.

[[[ incluir aqui parte da discussao sobre as opcoes de modelagem  com o argumento pela escolha dos subsets para a tipagem ]]] 
As our focus was on the interfaces of the Java Card API methods,

>>>>

\begin{document}
\begin{frontmatter}

  \title{Modeling the Java Card API in B}
  \author[ufrn]{A. Martins Moreira\thanksref{ALL}\thanksref{anamaria}}
  \author[ufrn]{S. de Oliveira Santos \thanksref{simone}}
  \author[ifrn]{B. Gurgel Gomes \thanksref{bruno}} \author[ufrn]{D.
    D\'{e}harbe \thanksref{david}} \address[ufrn]{DIMAp-UFRN, Natal,
    RN, Brazil} \thanks[ALL]{This work is partly supported by CAPES
    and CNPq grants 306033/2009-7, 560014/2010-4 and 573964/2008-4
    (National Institute of Science and Technology for Software
    Engineering---INES, \texttt{www.ines.org.br}).}
\address[ifrn]{IFRN, Currais Novos, RN, Brazil}
\thanks[anamaria]{Email: \href{mailto:anamaria@dimap.ufrn.br}
  {\texttt{\normalshape anamaria@dimap.ufrn.br}}}
\thanks[simone]{Email: \href{mailto:simone82@ppgsc.ufrn.br}
  {\texttt{\normalshape simone82@ppgsc.ufrn.br}}}
\thanks[bruno]{Email: \href{mailto:bruno.gurgel@ifrn.edu.br}
  {\texttt{\normalshape bruno.gurgel@ifrn.edu.br}}}
\thanks[david]{Email: \href{mailto:david@dimap.ufrn.br}
  {\texttt{\normalshape david@dimap.ufrn.br}}}

@@

Among the different components of the Java Card platform, the Java
Card API is the one without which no refinement into Java Card code
could be done.  Suppose, for instance, that we had the B abstract
machine of Fig.~\ref{lis:buscard}, which models a bus card with a
variable that keeps track of the current credit the user has
(\texttt{balance}) and an operation to add credits to this balance
(\texttt{add}). This simple example provides enough evidence of the
importance of having a B model of the Java Card API.  B tools can directly 
generate code for it in C, for instance, because it  is a deterministic specification and all data in
it is implementable in C.  The
resulting code would be very similar to the original B machine.

\lstset{language=B, caption={B machine Buscard},label=lis:buscard}
\begin{lstlisting}
MACHINE
	Buscard
CONSTANTS
	MAX_BALANCE
PROPERTIES
	MAX_BALANCE : INTEGER & MAX_BALANCE  = 100
VARIABLES
	balance
INVARIANT
	balance : INTEGER & balance <= MAX_BALANCE 
INITIALISATION
	balance := 0
OPERATIONS
	add(credit) =
	PRE
		 credit : INTEGER & 
		 credit > 0  & credit <= MAX_BALANCE &
                balance + credit <= MAX_BALANCE 
	THEN
		balance := balance + credit
	END
END
\end{lstlisting}

@@

\lstset{language=Java, caption={Method add in applet Java Card},label=lis:addjavacard}
\begin{lstlisting}
public void add (APDU apdu) { 
	byte[] bf;  byte cr; byte apdu_state; short res;

@@

		if (!(cr <= MAX_BALANCE)) {
			ISOException.throwIt(EX_CREDIT_EXCEEDED);
		} else if (!((short) (balance + cr) <= MAX_BALANCE)) {
			ISOException.throwIt(EX_BALANCE_EXCEEDED);
		} else if (! (cr > 0)) {
			ISOException.throwIt(EX_NEGATIVE_CREDIT);
		} else {
			balance = (short) (balance + cr);
		}
	} 
}
\end{lstlisting}

@@

		IF not (cr <= MAX_BALANCE)
		THEN 
			ISOException.throwIt(EX_CREDIT_EXCEEDED)
		ELSIF not (sum_short (balance, cr) <= MAX_BALANCE)
		THEN 
			ISOException.throwIt(EX_BALANCE_EXCEEDED)
		ELSIF not (cr > 0)
		THEN 
			ISOException.throwIt(EX_NEGATIVE_CREDIT) 
		ELSE  
			balance := sum_short(balance, cr)
		END
	END
END
\end{lstlisting}

@@

The change of interface is also taken care of by the BSmart method, so
that the development can be done without need to extend the B
method, as presented in~\cite{Gomes10}.

But all of these transformations rely on the need to reference Java
Card API types, classes and methods on the B level.  This is then the
reason why the Java Card API needed to be the first Java Card
component to be modeled in B. This is a common choice, as we can see in previous work on the formalization of 
Java Card in JML \cite{pollBerg01} and OCL \cite{Larsson:2004}.

@@

  \item In the current version of the specification, states are not modeled. 
  The B machines corresponding to Java Card classes or interfaces are stateless machines.

@@

\item The hierarchy of the Java Card API objects is partially modeled
  by abstract set inclusion. For instance, the class Throwable extends
  the class Object, in Java Card. In our model, we  have that the
  machine Throwable SEES the machine Object, and too show that all throwables are indeed
  objects, a property stating that THROWABLE is a subset of OBJECT is
  included, associated to the Throwable machine.

@@

As previously stated, we chose to model exception raising conditions
as operation pre-conditions in the B models. For instance, if the Java
Card specification states that a method should throw an exception in
case a parameter array is empty, the pre-condition of the
corresponding B operation will require that this parameter is not
empty. The B method then requires that every call to that operation
satisfies the operation's pre-condition. As a consequence, code that
is formally generated using the B method and the Java Card API
specification should never raise this specific exception, as it is
statically proved that all calls to this operation are made with
non-empty parameter arrays. Not all situations leading to exceptional
conditions are modeled, however, as some of them depend on the Java
Card Runtime Environment (JCRE) and its state, which has not yet been
modeled, and some require a more detailed specification of the
contents of some data and other internal behavior. In these cases, the
exception raising conditions are only left as comments to the B
specification, and can either be used as additional documentation or be
considered in future work.

One delicate issue to be dealt with both by the user of the API
specification and the tools designed to work with it, is the use of
overloaded methods in an application development. When overloaded
operations are used in an application development, the user must
explicitly choose which version (in B) of the overloaded operation is to be
used. Then, verifications will be carried out considering this
specific version of the operation. When code is generated, either the
code generator will remove the name extension automatically, if
designed for it, or the user will have to edit the generated code
manually. In any case, it is important that the actual method called
in the Java Card code is the one with which verifications have been
carried out.  In case of overriding, in particular, it will be either
the user's responsibility to choose the B one which is in scope at
this specific point in Java, or an extra verification step will be
included in the code generator to check that the operation used in
B is the one in scope in Java and registers an error in case it is
not.

Finally, the choices related to class derived types and return values were consequences of our focus on the interfaces of the Java Card API methods. The declaration of an abstract set for each class allows to model in B that a parameter of a particular class is expected or is to be returned by a method.

\lstset{language=B, caption={B machine Buscard},label=lis:object}
\begin{lstlisting}
MACHINE Object
SEES  ObjectSet
OPERATIONS
 	result <-- equals(obj) =
	PRE obj : OBJECT   /* declared in the ObjectSet machine */
	THEN
		result :: {TRUE, FALSE}
	END
END

MACHINE Throwable
/* This class extends java.lang.Object */
SEES
    Object,
    ThrowableSet
END

MACHINE ObjectSet    
SETS
     OBJECT   
END

MACHINE ThrowableSet
SEES 
    ObjectSet    
CONSTANTS  THROWABLE
PROPERTIES THROWABLE <: OBJECT
END
\end{lstlisting}
